---
title: "ESM 244 Lab 1"
author: "Katelyn Toigo"
date: "1/6/2022"
output: html_document
---

```{r setup, include=TRUE, warning = FALSE, message = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)

library(tidyverse)
library(here)
library(sf)
library(tmap)
### install.packages('tmap')
### update.packages(ask = FALSE)
```
 
 cmd-shift-enter is shortcut for running current code chunk
 
### Read in the data

cmd-option-i is shortcut for creating a code chunk

```{r}
sf_trees <- read_csv(here('data', 'sf_trees', 'sf_trees.csv'),
                     show_col_types = FALSE)
```

can put in console: comment out if in code.
View(sf_trees) shows the table of data
Name(sf_trees) shows the column names
summary(sf_trees) shows some stats and info

Hashtag outside of code chunk is the size of the header
Hashtag in a code chunk omit that note from the running code - comment it out

### Part 1: wrangling and ggplot review

**Example 1:** Find counts of observations by `legal_status` and wrangle a bit. 
2 asterisks before and after make it bold

```{r}
### method 1: group_by() %>% summarize()
sf_trees %>% # not assigned to an object, so will print to the page
  group_by(legal_status) %>% 
  summarize(tree_count = n()) # tree count is assigned to be the name of the rows

### method 2: different way plus a few new functions
top_5_status <- sf_trees %>% # assigned the function to an object this time
  count(legal_status) %>% 
  drop_na(legal_status) %>% 
  rename(tree_count = n) %>% 
  relocate(tree_count) %>% # makes the tree_count column the first on the table
  slice_max(tree_count, n=5) %>% 
  arrange(desc(tree_count)) # arrange(tree_count) puts them in order from low to high, if you want the opposite (high to low) then do either arrange(-tree_count) or arrange(desc(tree_count))
```

If you don't know what a function is doing, go to console and type `?drop_na` for example and it will define what the drop_na function does



Make a graph of top 5 from above:

```{r}
ggplot(data = top_5_status, aes(x = fct_reorder(legal_status,tree_count), y = tree_count)) +
  geom_col(fill = 'darkgreen') +
  labs(x = 'Legal status', y = 'Tree count') +
  coord_flip() +
  theme_minimal()
```



**Example 2:** Only going to keep observations where legal status is "Permitted Site" and caretaker is "MTA", and store as `permitted_data_df`

shift-cmd-c to comment/uncomment quickly

```{r}
# sf_trees$legal_status %>% unique()     This will tell the unique legal_status types if ran.
# unique(sf_trees$caretaker)

permitted_data_df <- sf_trees %>% 
  filter(legal_status == 'Permitted Site' & caretaker == 'MTA') # using a , or an & does the same thing. They say they want both to be true. If you use a vertical bar | then you are saying you want: one or the other or both at the same time
# if you want to say you want the legal status to be either permitted site or private then you could say: filter(legal_status %in% c('Permitted Site', 'Private') & caretaker == 'MTA')
```



**Example 3:** Only keep Blackwood Acacia trees, and then only keep columns `legal_status`, `date`, `latitude`, `longitude` and store as `blackwood_acacia_df`

```{r}
blackwood_acacia_df <- sf_trees %>%
  filter(str_detect(species, 'Blackwood Acacia')) %>% 
  select(legal_status, date, lat = latitude, lon = longitude)

### Make a little graph of locations
ggplot(data = blackwood_acacia_df, aes(x = lon, y = lat)) +
  geom_point(color = 'darkgreen')
```

**Example 4:** use `tidyr::separate()`

```{r}
sf_trees_sep <- sf_trees %>% 
  separate(species, into = c('spp_scientific', 'spp_common'), sep = ' :: ')
```

** Example 5:** use `tidyr::unite()`

```{r}
ex_5 <- sf_trees %>% 
  unite('id_status', tree_id, legal_status, sep = '_COOL_')
```






